\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2022-01-26 submission/manuscript.tex

\emph{Clostridioides difficile} colonization can result in no symptoms
or develop into an infection, ranging in severity from mild diarrhea to
toxic megacolon, sepsis, and death. Models that predict severity and
guide treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, it has its effect on CDI disease severity is
unclear and has not been incorporated into disease severity models. We
demonstrated that variation in the microbiome of mice colonized with
human feces yielded a range of disease outcomes. These results revealed
groups of bacteria associated with both severe and mild \emph{C.
difficile} infection outcomes. Gut bacterial community data from
patients with CDI could improve our ability to identify patients at risk
of developing more severe disease and improve interventions which target
\emph{C. difficile} and the gut bacteria to reduce host damage.

@@

Recently, Nagao-Kitamoto \emph{et al.} showed germ-free mice which
received a microbial community from different human donors resulted in
variation in a subsequent \emph{C. difficile} infection (14). Here, we
sought to further elucidate the relationship between members of the gut
bacterial community and CDI disease severity. We hypothesized that since
specific groups of gut bacteria affect the metabolism of \emph{C.
difficile} and its infection dynamics, we can also identify groups of
bacteria that affect the disease severity of the infection. To test this
hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
to create varied gut communities. We then challenged the mice with
\emph{C. difficile} ribotype 027 (RT027) and followed the mice for the
development of severe outcomes of moribundity and histopathologic cecal
tissue damage. Since the murine host and \emph{C. difficile} isolate
were the same and only the gut community varied, the variation in
disease severity we observed was attributable to the gut microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (15). We inoculated germ-free C57BL/6 mice with homogenized feces
from one of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (16). The gut communities were
allowed to equilibrate for two weeks post-inoculation (17). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (18, 19). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, such as recent
hospitalization or antibiotic use (16), we decided to test if \emph{C.
difficile} was able to infect these mice without any antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with a
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate spores. The mice were followed for 10
days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N1, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N3, N4, and N5, were not colonized at detectable levels on
the day after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (\emph{P} = 0.003). However, when looking at
each group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N2 and N5 through N9 had comparable toxin activity as the
moribund mice. Additionally, not all moribund mice had toxin activity
detected in their stool. Next, we examined the cecal tissue for
histopathologic damage. Moribund mice had high levels of epithelial
damage, tissue edema, and inflammation (Figure S1) similar to previously
reported histopathologic findings for \emph{C. difficile} RT027 (20). As
observed with toxin activity, the moribund mice had higher
histopathologic scores than the non-moribund mice (\emph{P} \textless{}
0.001). However, unlike the toxin activity, all moribund mice had
consistently high histopathologic summary scores (Figure 3B). The
non-moribund mice, Donor groups N1 through N9, had a range in tissue
damage from none detected to similar levels as the moribund mice, which
grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the donor
groups but were largely consistent within each donor group.

@@

Challenging mice colonized with different human fecal communities with
\emph{C. difficile} RT027 demonstrated that variation in members of the
gut microbiome affects \emph{C. difficile} infection disease severity.
Our analysis revealed an association between the relative abundance of
bacterial community members and disease severity. Previous studies
investigating the severity of CDI disease involving the microbiome have
had limited ability to interrogate this relationship between the
microbiome and disease severity. Studies that have used clinical data
have limited ability to control variation in the host, microbiome or
\emph{C. difficile} ribotype (21). Murine experiments typically use a
single mouse colony and different \emph{C. difficile} ribotypes to
create severity differences (22). Recently, our group has begun
uncovering the effect microbiome variation has on \emph{C. difficile}
infection. We showed the variation in the bacterial communities between
mice from different mouse colonies resulted in different clearance rates
of \emph{C. difficile} (23). We also showed varied ability of mice to
spontaneously eliminate \emph{C. difficile} infection when they were
treated with different antibiotics prior to \emph{C. difficile}
challenge (24). Overall, the results presented here have demonstrated
that the gut bacterial community contributed to the severity of \emph{C.
difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 25, 26). Those levels are driven by the
activities in the intestine (27). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (18, 28--31).
However, other studies have shown that disease is not necessarily linked
with toxin production (32) or the strain (33). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (34--37). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
22, 38, 39). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy gut community
and is only able to become established after a perturbation (19). Once
colonized, the different communities lead to different metabolic
responses and dynamics of the \emph{C. difficile} population (8, 24,
40). Gut bacteria metabolize primary bile acids into secondary bile
acids (41, 42). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (9, 10, 43,
44). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (45, 46). The nutrient environment
affects toxin production (47). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

A myriad of studies have explored the relationship between the
microbiome and CDI disease. CDI studies often use an infection model
with a single homogeneous bacterial community. Collins \emph{et al.}
used multiple human communities to colonize mice, however the
communities were pooled prior to gavaging into germ-free mice (48),
resulting in a single human-derived community. Studies examining
difference in disease often use different \emph{C. difficile} strains or
ribotypes in mice with similar microbiota as a proxy for variation in
disease, such as strain 630 for non-severe and RT027 for severe (18, 28,
29, 49). Studies have also demonstrated variation in infection through
tapering antibiotic dosage (19, 24, 50) or by reducing the amount of
\emph{C. difficile} cells or spores used for the challenge (18, 50).
These studies either lack variation in the initial microbiome or have
variation in the \emph{C. difficile} infection itself, confounding any
association between variation in severity and the microbiome.

Recently, Nagao-Kitamoto \emph{et al.} demonstrated when only allowing
the microbiome to vary in a murine \emph{C. difficile} infection
experiemnt that the resultant disease variation associated with the
microbiome communities (14). Our data have demonstrated gut bacterial
relative abundances associate with variation in toxin production,
histopathologic scoring of the cecal tissue and mortality. This analysis
revealed populations of \emph{Akkermansia}, \emph{Anaerostipes},
\emph{Coprobacillus}, \emph{Enterocloster}, \emph{Lactonifactor}, and
\emph{Monoglobus} were more abundant in the microbiome of non-moribund
mice which had low histopathologic scores and no detected toxin. The
protective role of these genera are supported by previous studies.
\emph{Coprobacillus}, \emph{Lactonifactor}, and \emph{Monoglobus} have
been shown to be involved in dietary fiber fermentation and associated
with healthy communities (51--54). \emph{Anaerostipes} and
\emph{Coprobacillus}, which produce short chain fatty acids, have been
associated with healthy communities (55--57). Furthermore,
\emph{Coprobacillus}, which was abundant in mice with low
histopathologic scores but rare in all other mice, has been shown to
contain a putative type I lantibiotic gene cluster and inhibit \emph{C.
difficile} colonization (58--60). \emph{Akkermansia} and
\emph{Enterocloster} were also identified as more abundant in mice which
had a low histopathologic scores but have contradictory supporting
evidence in the current literature. In our data, \emph{Akkermansia} was
most abundant in the non-moribund mice with low histopathologic scores
but there were some moribund mice which had increased populations of
\emph{Akkermansia}. This could be attributed to either a more protective
mucus layer was present inhibiting colonization (60, 61) or mucus
consumption by \emph{Akkermansia} could have been crossfeeding \emph{C.
difficile} or exposing a niche for \emph{C. difficile} (62--64).
Similarly, \emph{Enterocloster} was more abundant and associated with
low histopathologic scores. It has been associated with healthy
populations and has been used to mono-colonize germ-free mice to reduce
the ability of \emph{C. difficile} to colonize (65, 66). However,
\emph{Enterocloster} has also been involved in infections, such as
bacteremia (67, 68). These data have exemplified populations of bacteria
that have the potential to be either protective or harmful. Thus, the
disease outcome is not likely based on the abundance of individual
populations of bacteria, rather it is the result of the interactions of
the community.

The groups of bacteria that were associated with either a higher
histopathologic score or moribundity are members of the indigenous gut
community that also have been associated with disease, often referred to
as opportunistic pathogens. Many of the populations with pathogenic
potential that associated with worse outcomes are also facultative
anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
\emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
\emph{Streptococcus} have been shown to expand after antibiotic use (16,
69, 70) and are commonly detected in CDI cases (71--74). In addition to
these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
\emph{Helicobacter}, which associated with worse outcomes, have also
been associated with intestinal inflammation (75--77). Recently,
\emph{Helicobacter hepaticus} was shown to be sufficient to cause
susceptibility to CDI in IL-10 deficient C57BL/6 mice (78). In our
experiments, when \emph{Helicobacter} was present, the infection
resulted in a high histopathologic score (Figure 4C). While we did not
use IL-10 deficient mice, it is possible the bacterial community or host
response are similarly modified by \emph{Helicobacter}, allowing
\emph{C. difficile} infection and host damage. Aside from
\emph{Helicobacter}, these groups of bacteria that associated with more
severe outcomes did not have a conserved association between their
relative abundance and the disease severity across all mice.

Since we observed groups of bacteria that were associated with less
severe disease it may be appropriate to apply the damage-response
framework for microbial pathogenesis to CDI (79, 80). This framework
posits that disease is not driven by a single entity, rather it is an
emergent property of the responses of the host immune system, infecting
microbe, \emph{C. difficile}, and the indigenous microbes at the site of
infection. In the first set of experiments, we used the same host
background, C57BL/6 mice, the same infecting microbe, \emph{C.
difficile} RT027 clinical isolate 431, with different gut bacterial
communities. The bacterial groups in those communities were often
present in both moribund and non-moribund and across the range of
histopathologic scores. Thus, it was not merely the presence of the
bacteria but their activity in response to the other microbes and host
which affect the extent of the host damage. Additionally, while each
mouse and \emph{C. difficile} population had the same genetic
background, they too were reacting to the specific microbial community.
Disease severity is driven by the cumulative effect of the host immune
response and the activity of \emph{C. difficile} and the gut bacteria.
\emph{C. difficile} drives host damage through the production of toxin.
The gut microbiota can modulate host damage through the balance of
metabolic and competitive interactions with \emph{C. difficile}, such as
bacteriocin production or mucin degradation, and interactions with the
host, such as host mucus glycosylation or intestinal IL-33 expression
(14, 81). For example, low levels of mucin degradation can provide
nutrients to other community members producing a diverse non-damaging
community (82). However, if mucin degradation becomes too great it
reduces the protective function of the mucin layer and exposes the
epithelial cells. This over-harvesting can contribute to the host damage
due to other members producing toxin. Thus, the resultant intestinal
damage is the balance of all activities in the gut environment. Host
damage is the emergent property of numerous damage-response curves, such
as one for host immune response, one for \emph{C. difficile} activity
and another for microbiome community activity, each of which are a
composite curve of the individual activities from each group, such as
antibody production, neutrophil infiltration, toxin production,
sporulation, fiber and mucin degradation. Therefore, while we have
identified populations of interest, it may be necessary to target
multiple types of bacteria to reduce the community interactions
contributing to host damage.

@@

\textbf{\emph{C. difficile} experiments.} Human fecal samples were
obtained as part of Schubert \emph{et al.} and selected based on
community clusters (16) to result in diverse community structures. Feces
were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
inoculated with 100 \(\mu\)l of the fecal homogenate via oral gavage.
Two weeks after the fecal community inoculation, mice were challenged
with \emph{C. difficile}. \emph{C. difficile} clinical isolate 431 came
from Carlson \emph{et al.} which had previously been isolated and
characterized (34, 35) and has recently been further characterized (36).
Spores concentration were determined both before and after challenge
(83). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
oral gavage.

\textbf{Sample collection.} Fecal samples were collected on the day of
\emph{C. difficile} challenge and the following 10 days. Each day, a
fecal sample was collected and a portion was weighed for plating
(approximately 30 mg) and the remaining sample was frozen at
-20\(^\circ\)C. Anaerobically, the weighed fecal samples were serially
diluted in PBS, plated on TCCFA plates, and incubated at 37\(^\circ\)C
for 24 hours. The plates were then counted for the number of colony
forming units (CFU) (84).

\textbf{DNA sequencing.} From the frozen fecal samples, total bacterial
DNA was extracted using MOBIO PowerSoil-htp 96-well soil DNA isolation
kit. We amplified the 16S rRNA gene V4 region and sequenced the
resulting amplicons using an Illumina MiSeq as described previously
(85).

\textbf{Sequence curation.} Sequences were processed with
mothur(v.1.44.3) as previously described (85, 86). In short, we used a
3\% dissimilarity cutoff to group sequences into operational taxonomic
units (OTUs). We used a naive Bayesian classifier with the Ribosomal
Database Project training set (version 18) to assign taxonomic
classifications to each OTU (87). We sequenced a mock community of a
known community composition and 16s rRNA gene sequences. We processed
this mock community with our samples to calculate the error rate for our
sequence curation, which was an error rate of 0.19\%.

@@

\leavevmode\hypertarget{ref-Nagpal2018}{}%
15. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
\textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
\textbf{Register TC}, \textbf{Craft S}, \textbf{McClain DA},
\textbf{Yadav H}. 2018. Comparative microbiome signatures and
short-chain fatty acids in mouse, rat, non-human primate, and human
feces. Frontiers in Microbiology \textbf{9}.
doi:\href{https://doi.org/10.3389/fmicb.2018.02897}{10.3389/fmicb.2018.02897}.

>>>>

\emph{Clostridioides difficile} colonization can be asymptomatic or
develop into an infection, ranging in severity from mild diarrhea to
toxic megacolon, sepsis, and death. Models that predict severity and
guide treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, its effect on CDI disease severity is unclear
and has not been incorporated into disease severity models. We
demonstrated that variation in the microbiome of mice colonized with
human feces yielded a range of disease outcomes. These results revealed
groups of bacteria associated with both severe and mild \emph{C.
difficile} infection outcomes. Gut bacterial community data from
patients with CDI could improve our ability to identify patients at risk
of developing more severe disease and improve interventions which target
\emph{C. difficile} and the gut bacteria to reduce host damage.

@@

Recent studies have demonstrated that when mice with diverse microbial
communities were challenged with a high-toxigenic strain resulted in
varied disease severity (14) and when challenged with a low-toxigenic
strain members of the gut microbial community associated with variation
in colonization (15). Here, we sought to further elucidate the
relationship between members of the gut bacterial community and CDI
disease severity when challenged with a high-toxigenic strain, \emph{C.
difficile} ribotype 027 (RT027). We hypothesized that since specific
groups of gut bacteria affect the metabolism of \emph{C. difficile} and
its infection dynamics, we can also identify groups of bacteria that
affect the disease severity of the infection. To test this hypothesis,
we colonized germ-free C57BL/6 mice with human fecal samples to create
varied gut communities. We then challenged the mice with \emph{C.
difficile} RT027 and followed the mice for the development of severe
outcomes of moribundity and histopathologic cecal tissue damage. Since
the murine host and \emph{C. difficile} isolate were the same and only
the gut community varied, the variation in disease severity we observed
was attributable to the gut microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics.} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (16). We inoculated germ-free C57BL/6 mice with homogenized feces
from each of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (17). The gut communities were
allowed to equilibrate for two weeks post-inoculation (18). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (19, 20). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, such as recent
hospitalization or antibiotic use (17), we tested whether \emph{C.
difficile} was able to infect these mice without an antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with a
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate spores. The mice were followed for 10
days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N1, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N3, N4, and N5, were not colonized at detectable levels on
the day after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (\emph{P} = 0.003). However, when looking at
each group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N2 and N5 through N9 had comparable toxin activity as the
moribund mice. Additionally, not all moribund mice had toxin activity
detected in their stool. Next, we examined the cecal tissue for
histopathologic damage. Moribund mice had high levels of epithelial
damage, tissue edema, and inflammation (Figure S1) similar to previously
reported histopathologic findings for \emph{C. difficile} RT027 (21). As
observed with toxin activity, the moribund mice had higher
histopathologic scores than the non-moribund mice (\emph{P} \textless{}
0.001). However, unlike the toxin activity, all moribund mice had
consistently high histopathologic summary scores (Figure 3B). The
non-moribund mice, Donor groups N1 through N9, had a range in tissue
damage from none detected to similar levels as the moribund mice, which
grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the donor
groups but were largely consistent within each donor group.

@@

Challenging mice colonized with different human fecal communities with
\emph{C. difficile} RT027 demonstrated that variation in members of the
gut microbiome affects \emph{C. difficile} infection disease severity.
Our analysis revealed an association between the relative abundance of
bacterial community members and disease severity. Previous studies
investigating the severity of CDI disease involving the microbiome have
had limited ability to interrogate this relationship between the
microbiome and disease severity. Studies that have used clinical data
have limited ability to control variation in the host, microbiome or
\emph{C. difficile} ribotype (22). Murine experiments typically use a
single mouse colony and different \emph{C. difficile} ribotypes to
create severity differences (23). Recently, our group has begun
uncovering the effect microbiome variation has on \emph{C. difficile}
infection. We showed the variation in the bacterial communities between
mice from different mouse colonies resulted in different clearance rates
of \emph{C. difficile} (15). We also showed varied ability of mice to
spontaneously eliminate \emph{C. difficile} infection when they were
treated with different antibiotics prior to \emph{C. difficile}
challenge (24). Overall, the results presented here have demonstrated
that the gut bacterial community contributed to the severity of \emph{C.
difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 25, 26). Those levels are driven by the
activities in the intestine (27). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (19, 28--31).
However, other studies have shown that disease is not necessarily linked
with toxin production (32) or the strain (33). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (34--37). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
23, 38, 39). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy murine gut
community and is only able to become established after a perturbation
(20). Once colonized, the different communities lead to different
metabolic responses and dynamics of the \emph{C. difficile} population
(8, 24, 40). Gut bacteria metabolize primary bile acids into secondary
bile acids (41, 42). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (9, 10, 43,
44). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (45, 46). The nutrient environment
affects toxin production (47). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

A myriad of studies have explored the relationship between the
microbiome and CDI disease. Studies examining difference in disease
often use different \emph{C. difficile} strains or ribotypes in mice
with similar microbiota as a proxy for variation in disease, such as
strain 630 for non-severe and RT027 for severe (19, 28, 29, 48). Studies
have also demonstrated variation in infection through tapering
antibiotic dosage (20, 24, 49) or by reducing the amount of \emph{C.
difficile} cells or spores used for the challenge (19, 49). These
studies often either lack variation in the initial microbiome or have
variation in the \emph{C. difficile} infection itself, confounding any
association between variation in severity and the microbiome. Recent
studies have shown variation in the initial microbiome, via different
murine colonies or colonizing germ-free mice with human feces, that were
challenged with \emph{C. difficile} resulted in varied outcomes of the
infection (14, 15).

Our data have demonstrated gut bacterial relative abundances associate
with variation in toxin production, histopathologic scoring of the cecal
tissue and mortality. This analysis revealed populations of
\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
more abundant in the microbiome of non-moribund mice which had low
histopathologic scores and no detected toxin. The protective role of
these genera are supported by previous studies. \emph{Coprobacillus},
\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
involved in dietary fiber fermentation and associated with healthy
communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
which produce short chain fatty acids, have been associated with healthy
communities (54--56). Furthermore, \emph{Coprobacillus}, which was
abundant in mice with low histopathologic scores but rare in all other
mice, has been shown to contain a putative type I lantibiotic gene
cluster and inhibit \emph{C. difficile} colonization (57--59).
\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
abundant in mice which had a low histopathologic scores but have
contradictory supporting evidence in the current literature. In our
data, \emph{Akkermansia} was most abundant in the non-moribund mice with
low histopathologic scores but there were some moribund mice which had
increased populations of \emph{Akkermansia}. This could be attributed to
either a more protective mucus layer was present inhibiting colonization
(59, 60) or mucus consumption by \emph{Akkermansia} could have been
crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
and associated with low histopathologic scores. It has been associated
with healthy populations and has been used to mono-colonize germ-free
mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
However, \emph{Enterocloster} has also been involved in infections, such
as bacteremia (66, 67). These data have exemplified populations of
bacteria that have the potential to be either protective or harmful.
Thus, the disease outcome is not likely based on the abundance of
individual populations of bacteria, rather it is the result of the
interactions of the community.

The groups of bacteria that were associated with either a higher
histopathologic score or moribundity are members of the indigenous gut
community that also have been associated with disease, often referred to
as opportunistic pathogens. Many of the populations with pathogenic
potential that associated with worse outcomes are also facultative
anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
\emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
\emph{Streptococcus} have been shown to expand after antibiotic use (17,
68, 69) and are commonly detected in CDI cases (70--73). In addition to
these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
\emph{Helicobacter}, which associated with worse outcomes, have also
been associated with intestinal inflammation (74--76). Recently,
\emph{Helicobacter hepaticus} was shown to be sufficient to cause
susceptibility to CDI in IL-10 deficient C57BL/6 mice (77). In our
experiments, when \emph{Helicobacter} was present, the infection
resulted in a high histopathologic score (Figure 4C). While we did not
use IL-10 deficient mice, it is possible the bacterial community or host
response are similarly modified by \emph{Helicobacter}, allowing
\emph{C. difficile} infection and host damage. Aside from
\emph{Helicobacter}, these groups of bacteria that associated with more
severe outcomes did not have a conserved association between their
relative abundance and the disease severity across all mice.

Since we observed groups of bacteria that were associated with less
severe disease it may be appropriate to apply the damage-response
framework for microbial pathogenesis to CDI (78, 79). This framework
posits that disease is not driven by a single entity, rather it is an
emergent property of the responses of the host immune system, infecting
microbe, \emph{C. difficile}, and the indigenous microbes at the site of
infection. In the first set of experiments, we used the same host
background, C57BL/6 mice, the same infecting microbe, \emph{C.
difficile} RT027 clinical isolate 431, with different gut bacterial
communities. The bacterial groups in those communities were often
present in both moribund and non-moribund and across the range of
histopathologic scores. Thus, it was not merely the presence of the
bacteria but their activity in response to the other microbes and host
which affect the extent of the host damage. Additionally, while each
mouse and \emph{C. difficile} population had the same genetic
background, they too were reacting to the specific microbial community.
Disease severity is driven by the cumulative effect of the host immune
response and the activity of \emph{C. difficile} and the gut bacteria.
\emph{C. difficile} drives host damage through the production of toxin.
The gut microbiota can modulate host damage through the balance of
metabolic and competitive interactions with \emph{C. difficile}, such as
bacteriocin production or mucin degradation, and interactions with the
host, such as host mucus glycosylation or intestinal IL-33 expression
(14, 80). For example, low levels of mucin degradation can provide
nutrients to other community members producing a diverse non-damaging
community (81). However, if mucin degradation becomes too great it
reduces the protective function of the mucin layer and exposes the
epithelial cells. This over-harvesting can contribute to the host damage
due to other members producing toxin. Thus, the resultant intestinal
damage is the balance of all activities in the gut environment. Host
damage is the emergent property of numerous damage-response curves, such
as one for host immune response, one for \emph{C. difficile} activity
and another for microbiome community activity, each of which are a
composite curve of the individual activities from each group, such as
antibody production, neutrophil infiltration, toxin production,
sporulation, fiber and mucin degradation. Therefore, while we have
identified populations of interest, it may be necessary to target
multiple types of bacteria to reduce the community interactions
contributing to host damage.

@@

\textbf{\emph{C. difficile} experiments.} Human fecal samples were
obtained as part of Schubert \emph{et al.} and selected based on
community clusters (17) to result in diverse community structures. Feces
were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
inoculated with 100 \(\mu\)l of the fecal homogenate via oral gavage.
Two weeks after the fecal community inoculation, mice were challenged
with \emph{C. difficile}. \emph{C. difficile} clinical isolate 431 came
from Carlson \emph{et al.} which had previously been isolated and
characterized (34, 35) and has recently been further characterized (36).
Spores concentration were determined both before and after challenge
(82). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
oral gavage.

\textbf{Sample collection.} Fecal samples were collected on the day of
\emph{C. difficile} challenge and the following 10 days. Each day, a
fecal sample was collected and a portion was weighed for plating
(approximately 30 mg) and the remaining sample was frozen at
-20\(^\circ\)C. Anaerobically, the weighed fecal samples were serially
diluted in PBS, plated on TCCFA plates, and incubated at 37\(^\circ\)C
for 24 hours. The plates were then counted for the number of colony
forming units (CFU) (83).

\textbf{DNA sequencing.} From the frozen fecal samples, total bacterial
DNA was extracted using MOBIO PowerSoil-htp 96-well soil DNA isolation
kit. We amplified the 16S rRNA gene V4 region and sequenced the
resulting amplicons using an Illumina MiSeq as described previously
(84).

\textbf{Sequence curation.} Sequences were processed with
mothur(v.1.44.3) as previously described (84, 85). In short, we used a
3\% dissimilarity cutoff to group sequences into operational taxonomic
units (OTUs). We used a naive Bayesian classifier with the Ribosomal
Database Project training set (version 18) to assign taxonomic
classifications to each OTU (86). We sequenced a mock community of a
known community composition and 16s rRNA gene sequences. We processed
this mock community with our samples to calculate the error rate for our
sequence curation, which was an error rate of 0.19\%.

@@

\leavevmode\hypertarget{ref-Tomkovich2020}{}%
15. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
clearance in mice. mSphere \textbf{5}.
doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
